Complex Synapse Identification

### 2014-07-04 Notes/Identification.tex
notation changes

Let $t$ be the time of a plasticity event, assumed to be an integer. 
We denote a sequence of event times by $[0,T]$.

We denote states of a synapse by subscripts $i,j$. 
The state at a particular time during a sequence is $i(t)$.
An entire trajectory (sequence of states) is denoted by $i[0,T]$.

@@

The different types of plasticity (\eg potentiation and depression) are indicated by superscripts $\mu,\nu$.
The type of plasticity used at time $t$ is $\mu(t)$.
An entire sequence of plasticity types is denoted by $\mu[0,T]$.

@@

When we have several sequences, we can write
%
\begin{multline}\label{eq:emqseveral}
%\begin{aligned}
  Q(\M,\M') = \sum_{i^A[0,T]} \brk{ \prod_a \pr\cond{i^a[0,T],\w^a[0,T]}{\M} } \brk{\sum_a  \log\!\!  \brk{
      \frac{ \pr\cond{i^A[0,T],\w^a[0,T]}{\M'} }{ \pr\cond{i^A[0,T],\w^a[0,T]}{\M} } } }\\
   = \sum_a \prod_{b \neq a} \pr\cond{\w^b[0,T]}{\M} \sum_{i^a[0,T]} \pr\cond{i^a[0,T],\w^a[0,T]}{\M} \log\!\!  \brk{
      \frac{ \pr\cond{i^a[0,T],\w^a[0,T]}{\M'} }{ \pr\cond{i^a[0,T],\w^a[0,T]}{\M} } } \\
   = \pr\cond{\w^A[0,T]}{\M} \sum_a \frac{Q^a(\M,\M')}{\pr\cond{\w^a[0,T]}{\M}}.
%\end{aligned}
\end{multline}
%

@@

Suppose we have a single sequence of plasticity events.
Then the joint likelihood of a trajectory and sequence observed synaptic weights is given by
%
\begin{equation}\label{eq:trajlike}
\begin{aligned}
  \pr\cond{t[0,T], \w[0,T]}{\M} &= \pib_{i(0)} \, \Pib(0) 
        \prod_{t=0}^{T-1} \M^{\mu(t)}_{i(t)i(t+1)} \, \Pib(t+1),\\
  \text{where}\quad
  \Pib(t) &= \delta_{\w_{i(t)}\w(t)}
\end{aligned}
\end{equation}
%
We wish to maximise \eqref{eq:emq} subject to normalisation constraints.
Therefore we maximise the Lagrangian:
%
\begin{equation}\label{eq:lagrangesingle}
  \CL = Q(\M,\M') + \lambda\prn{1-\sum_i\pib'_i} + \sum_{\mu i} \lambda^\mu_i\prn{1-\sum_j\M'^\mu_{ij}}. 
\end{equation}
%
We find
%
\begin{equation}\label{eq:singlediffs}
\begin{aligned}
  \pdiff{\CL}{\M'^\mu_{ij}} &= \frac{N^\mu_{ij}}{\M'^\mu_{ij}} - \lambda^\mu_{i}, \\
  \pdiff{\CL}{\pib'_{i}} &= \frac{\alpha_i(0) \, \beta_i(0)}{\pib'_{i}} - \lambda,
\end{aligned}
\end{equation}
%
where the numerator $N^\mu_{ij}$ and the forward/backward variables $\alpha$ and $\beta$ are defined as:
%
\begin{equation}\label{eq:alphabeta}
\begin{aligned}
  N^\nu_{i(t)i(t+1)} &= \sum_{t=0}^{T-1} \pr\cond{i[t,t+1]),\w[0,T]}{\M} \\
     &= \sum_{t=0}^{T-1} \delta_{\mu(t)\nu} \, \alpha_i(t) \, \beta_j(t+1) \, \M^{\nu}_{i(t)i(t+1)} \, \Pib(t+1) ,\\
  \alpha_{i(t)}(t) &= \pr\cond{i(t),\w[0,t]}{\M} \\
     &= \sum_{i[0,t-1]} \pib_{i(0)} \, \delta_{\w_{i(0)}\w(0)} \prod_{s=0}^{t-1} \M^{\mu(s)}_{i(s)i(s+1)} \, \Pib(s+1), \\
  \beta_{i(t)}(t) &= \pr\cond{i(t),\w[t+1,T]}{\M} \\
     &= \sum_{i[t+1,T]} \prod_{s=t}^{T-1} \M^{\mu(s)}_{i(s)i(s+1)} \, \Pib(s+1).
\end{aligned}
\end{equation}
%
Note that, whilst disjoint sequences $\w[s,s']$ and $\w[t,t']$ are not independent, they are conditionally independent given the state at any time between the two intervals.
Therefore:
%
\begin{equation}\label{eq:stateprob}
  \pr\cond{i(t),\w[0,T]}{\M} = \alpha_{i(t)}(t) \, \beta_{i(t)}(t).
\end{equation}
%

These quantities usually suffer from underflow.
This can be avoided by using normalised forwar/backward variables \cite{Zhai2003HMMnorm}:
%
\begin{equation}\label{eq:normalphabeta}
\begin{aligned}
  \tilde{\alpha}_i(t) &= \alpha_i(t) \prod_{s=0}^{t} \eta(s), \\
  \tilde{\beta}_i(t) &= \beta_i(t) \prod_{s=t+1}^{T} \eta(s),
\end{aligned}
\end{equation}
%
where $\eta[0,T]$ is chosen so that all of the $\tilde{\alpha}[0,T]$ are normalised: $\sum_i \tilde{\alpha}_i(t) = 1$.
These quantities have the interpretations:
%
\begin{equation}\label{eq:normintepr}
\begin{aligned}
  \tilde{\alpha}_{i(t)}(t) &= \pr\cond{i(t)}{\w[0,t],\M}, \\
  \eta(t) &= \pr\cond{\w(t)}{\w[0,t-1],\M}, \\
  \prod_{s=0}^t \eta(t) &= \pr\cond{\w[0,t]}{\M}.
\end{aligned}
\end{equation}
%
Note that the new backward variables $\tilde{\beta}$ are not normalised and have no particularly interesting interpretation.
However:
%
\begin{equation}\label{eq:normstateprob}
\begin{aligned}
  \widetilde{N}^\nu_{i(t)i(t+1)} &= \sum_{t=0}^{T-1} \pr\cond{i[t,t+1])}{\w[0,T],\M} \\
     &= \sum_{t=0}^{T-1} \delta_{\mu(t)\nu} \, \tilde{\alpha}_i(t) \, \tilde{\beta}_j(t+1) \, \eta(t+1) \, \M^{\nu}_{i(t)i(t+1)} \, \Pib(t+1) ,\\
  \pr\cond{i(t)}{\w[0,T],\M} &= \alpha_{i(t)}(t) \, \beta_{i(t)}(t).
\end{aligned}
\end{equation}
%
Then we can write
%
\begin{equation}\label{eq:singlediffsnorm}
\begin{aligned}
  \pdiff{\CL}{\M'^\mu_{ij}} &= \pr\cond{\w[0,T]}{\M} \, \frac{ \widetilde{N}^\mu_{ij} }{ \M'^\mu_{ij} } - \lambda^\mu_{i}, \\
  \pdiff{\CL}{\pib'_{i}} &= \pr\cond{\w[0,T]}{\M} \, \frac{ \tilde{\alpha}_i(0) \, \tilde{\beta}_i(0) }{ \pib'_{i} } - \lambda,
\end{aligned}
\end{equation}
%
The factors of $\pr\cond{\w[0,T]}{\M}$ can be absorbed by a redefinition of the Lagrange multipliers.

@@

\subsection{Off-diagonal $\CL^1$ penalty \label{sec:offdiagL1}}

@@

\subsection{$\CL^\half$ penalty \label{sec:Lhalf}}

Here we use the  $\CL^\half$ norm as a penalty:
%
\begin{equation}\label{eq:Lhalfpenalty}
  E^\mu_{ij} = 2\sqrt{\M^\mu_{ij}}.
\end{equation}
%
Then the condition for a maximum \eqref{eq:priordiffs} reads
%
\begin{equation}\label{eq:Lhalfpriormax}
  \frac{ \widetilde{N}^{A\mu}_{ij} }{ \M'^\mu_{ij} }  - \frac{\beta}{\sqrt{\M'^\mu_{ij}}} - \lambda^\mu_{i} = 0.
\end{equation}
%
It is helpful to define
%
\begin{equation}\label{eq:Lhalfvars}
\begin{aligned}
  \widetilde{N}^{A\mu}_{ij} &= \beta^2 L^\mu_{ij}, \\
  \lambda^\mu_i &= \prn{\gamma^\mu_i}^{-2}.
\end{aligned}
\end{equation}
%
Then we have the update rule
%
\begin{equation}\label{eq:Lhalfupdate}
  \M'^\mu_{ij} = \frac{ \prn{\gamma^\mu_i}^2 \beta^2 }{ 4 } \prn{\sqrt{ \prn{\gamma^\mu_i}^2 + 4\prn{L^\mu_{ij}}^2 } - \gamma^\mu_i }^2, 
\end{equation}
%
with $\gamma^\mu_i$ determined by:
%
\begin{equation}\label{eq:Lhalfnorm}
    \frac{ \prn{\gamma^\mu_i}^2 \beta^2 }{ 4 } \sum_j \prn{\sqrt{\prn{\gamma^\mu_i}^2 + 4\prn{L^\mu_{ij}}^2} - \gamma^\mu_i}^2 = 1.
\end{equation}
%

>>>>

Let $t$ be the time of a plasticity event, assumed to be an integer. 
We denote a sequence of event times by $[0,T]$.
Any sequence whose upper bound is smaller that its lower bound can be considered empty, and that variable is omitted.
The same applies to sums and products.

We denote states of a synapse by subscripts $i,j,k$. 
The state at a particular time during a sequence is $i(t)$.
An entire trajectory (sequence of states) is denoted by $i[0,T]$.

@@

The different types of plasticity (\eg potentiation and depression) are indicated by superscripts $\mu,\nu$.
The type of plasticity used between times $(t-1,t)$ is $\mu(t)$.
An entire sequence of plasticity types is denoted by $\mu[1,T]$.

@@

When we have several sequences, we can write
%
\begin{equation}\label{eq:emqseveral}
\begin{gathered}
  Q(\M,\M') = \sum_{i^A[0,T]}  \prod_b \pr\cond{i^b[0,T],\w^b[0,T]}{\M} \sum_a  \log\!\!  \brk{
      \frac{ \pr\cond{i^a[0,T],\w^a[0,T]}{\M'} }{ \pr\cond{i^a[0,T],\w^a[0,T]}{\M} } } \\
   = \sum_a \prod_{b \neq a} \pr\cond{\w^b[0,T]}{\M}\! \sum_{i^a[0,T]}\!\! \pr\cond{i^a[0,T],\w^a[0,T]}{\M} \log\!\!  \brk{
      \frac{ \pr\cond{i^a[0,T],\w^a[0,T]}{\M'} }{ \pr\cond{i^a[0,T],\w^a[0,T]}{\M} } } \\
   = \pr\cond{\w^A[0,T]}{\M} \sum_a \frac{Q^a(\M,\M')}{\pr\cond{\w^a[0,T]}{\M}}.
\end{gathered}
\end{equation}
%

@@

Suppose we have a single sequence of plasticity events.
Then the joint likelihood of a trajectory and sequence observed synaptic weights is given by
%
\begin{equation}\label{eq:trajlike}
\begin{aligned}
  \pr\cond{t[0,T], \w[0,T]}{\M} &= \pib_{i(0)} \, \Pib_{i(0)}(0) 
        \prod_{t=1}^{T} \M^{\mu(t)}_{i(t-1)i(t)} \, \Pib_{i(t)}(t),\\
  \text{where}\quad
  \Pib_i(t) &= \delta_{\w_{i}\w(t)}
\end{aligned}
\end{equation}
%
We wish to maximise \eqref{eq:emq} subject to normalisation constraints.
Therefore we maximise the Lagrangian:
%
\begin{equation}\label{eq:lagrangesingle}
  \CL = Q(\M,\M') + \lambda\prn{1-\sum_i\pib'_i} + \sum_{\mu i} \lambda^\mu_i\prn{1-\sum_j\M'^\mu_{ij}}. 
\end{equation}
%
We find
%
\begin{equation}\label{eq:singlediffs}
\begin{aligned}
  \pdiff{\CL}{\M'^\mu_{ij}} &= \frac{N^\mu_{ij}}{\M'^\mu_{ij}} - \lambda^\mu_{i}, \\
  \pdiff{\CL}{\pib'_{i}} &= \frac{\alpha_i(0) \, \beta_i(0)}{\pib'_{i}} - \lambda,
\end{aligned}
\end{equation}
%
where the numerator $N^\mu_{ij}$ and the forward/backward variables $\alpha$ and $\beta$ are defined as:
%
\begin{equation}\label{eq:alphabeta}
\begin{aligned}
  N^\nu_{jk} &= \sum_{t=1}^{T} \delta_{\mu(t)\nu} \, \pr\cond{i[t-1,t]=(jk),\w[0,T]}{\M} \\
     &= \sum_{t=1}^{T} \delta_{\mu(t)\nu} \, \alpha_j(t-1) \, \beta_k(t) \, \M^{\nu}_{jk} \, \Pib_k(t) ,\\
  \alpha_j(t) &= \pr\cond{i(t)=j,\w[0,t]}{\M} \\
     &= \sum_{i[0,t]} \pib_{i(0)}(0) \, \Pib_{i(0)}(0) \prod_{s=1}^{t} \M^{\mu(s)}_{i(s-1)i(s)} \, \Pib_{i(s)}(s) \,\delta_{i(t)j}, \\
  \beta_j(t) &= \pr\cond{i(t)=j,\w[t+1,T]}{\M} \\
     &= \sum_{i[t,T]} \delta_{i(t)j} \prod_{s=t+1}^{T} \M^{\mu(s)}_{i(s-1)i(s)} \, \Pib_{i(s)}(s).
\end{aligned}
\end{equation}
%
Note that, whilst disjoint sequences $\w[s,s']$ and $\w[t,t']$ are not independent, they are conditionally independent given the state at any time between the two intervals.
Therefore:
%
\begin{equation}\label{eq:stateprob}
  \alpha_j(t) \, \beta_j(t) = \pr\cond{i(t)=j,\w[0,T]}{\M}.
\end{equation}
%

These quantities usually suffer from underflow.
This can be avoided by using normalised forward/backward variables \cite{Zhai2003HMMnorm}:
%
\begin{equation}\label{eq:normalphabeta}
\begin{aligned}
  \tilde{\alpha}_i(t) &= \alpha_i(t) \prod_{s=0}^{t} \eta(s), \\
  \tilde{\beta}_i(t) &= \beta_i(t) \prod_{s=t+1}^{T} \eta(s),
\end{aligned}
\end{equation}
%
where $\eta[0,T]$ is chosen so that all of the $\tilde{\alpha}[0,T]$ are normalised: $\sum_i \tilde{\alpha}_i(t) = 1$.
These quantities have the interpretations:
%
\begin{equation}\label{eq:normintepr}
\begin{aligned}
  \tilde{\alpha}_j(t) &= \pr\cond{i(t)=j}{\w[0,t],\M}, \\
  \eta(t) &= \pr\cond{\w(t)}{\w[0,t-1],\M}, \\
  \prod_{s=0}^t \eta(t) &= \pr\cond{\w[0,t]}{\M}.
\end{aligned}
\end{equation}
%
Note that the new backward variables $\tilde{\beta}$ are not normalised and have no particularly interesting interpretation.
However:
%
\begin{equation}\label{eq:normstateprob}
\begin{aligned}
  \widetilde{N}^\nu_{jk} &= \sum_{t=1}^{T} \delta_{\mu(t)\nu} \, \pr\cond{i[t-1,t]=(jk)}{\w[0,T],\M} \\
     &= \sum_{t=1}^{T} \delta_{\mu(t)\nu} \, \tilde{\alpha}_i(t-1) \, \tilde{\beta}_j(t) \, \eta(t) \, \M^{\nu}_{i(t-1)i(t)} \, \Pib_k(t) ,\\
  \tilde{\alpha}_j(t) \, \tilde{\beta}_j(t) &= \pr\cond{i(t)=j}{\w[0,T],\M}.
\end{aligned}
\end{equation}
%
Then we can write
%
\begin{equation}\label{eq:singlediffsnorm}
\begin{aligned}
  \pdiff{\CL}{\M'^\mu_{ij}} &= \pr\cond{\w[0,T]}{\M} \, \frac{ \widetilde{N}^\mu_{ij} }{ \M'^\mu_{ij} } - \lambda^\mu_{i}, \\
  \pdiff{\CL}{\pib'_{i}} &= \pr\cond{\w[0,T]}{\M} \, \frac{ \tilde{\alpha}_i(0) \, \tilde{\beta}_i(0) }{ \pib'_{i} } - \lambda,
\end{aligned}
\end{equation}
%
The factors of $\pr\cond{\w[0,T]}{\M}$ can be absorbed by a redefinition of the Lagrange multipliers.

@@

\subsection{Off-diagonal \texorpdfstring{$\CL^1$}{L\textasciicircum1} penalty \label{sec:offdiagL1}}

@@

\subsection{\texorpdfstring{$\CL^\half$}{L\textasciicircum1/2} penalty \label{sec:Lhalf}}

Here we use the  $\CL^\half$ norm as a penalty:
%
\begin{equation}\label{eq:Lhalfpenalty}
  E^\mu_{ij} = 2\sqrt{\M^\mu_{ij}}.
\end{equation}
%
Then the condition for a maximum \eqref{eq:priordiffs} reads
%
\begin{equation}\label{eq:Lhalfpriormax}
  \frac{ \widetilde{N}^{A\mu}_{ij} }{ \M'^\mu_{ij} }  - \frac{\beta}{\sqrt{\M'^\mu_{ij}}} - \lambda^\mu_{i} = 0.
\end{equation}
%
It is helpful to define
%
\begin{equation}\label{eq:Lhalfvars}
\begin{aligned}
  \widetilde{N}^{A\mu}_{ij} &= \beta^2 L^\mu_{ij}, \\
  \lambda^\mu_i &= \prn{\gamma^\mu_i}^{-2}.
\end{aligned}
\end{equation}
%
Then we have the update rule
%
\begin{equation}\label{eq:Lhalfupdate}
  \M'^\mu_{ij} = \frac{ \prn{\gamma^\mu_i}^2 \beta^2 }{ 4 } \prn{\sqrt{ \prn{\gamma^\mu_i}^2 + 4L^\mu_{ij} } - \gamma^\mu_i }^2,
\end{equation}
%
with $\gamma^\mu_i$ determined by:
%
\begin{equation}\label{eq:Lhalfnorm}
    \frac{ \prn{\gamma^\mu_i}^2 \beta^2 }{ 4 } \sum_j \prn{\sqrt{\prn{\gamma^\mu_i}^2 + 4L^\mu_{ij}} - \gamma^\mu_i}^2 = 1.
\end{equation}
%
